Using generalized additive models to analyze biomedical non-linear longitudinal data

### 2021-06-22 Manuscripts/Manuscript by chapters -for submission/Full_document_SAGE.tex
code highlighting

@@

\begin{abstract}
In biomedical research, the outcome of longitudinal studies has been
traditionally analyzed using the \emph{repeated measures analysis of
variance} (rm-ANOVA) or more recently, \emph{linear mixed models}
(LMEMs). Although LMEMs are less restrictive than rm-ANOVA in terms of
correlation and missing observations, both methodologies share an
assumption of linearity in the measured response, which results in
biased estimates and unreliable inference when they are used to analyze
data where the trends are non-linear, which is a common occurrence in
biomedical research. In contrast, \emph{generalized additive models}
(GAMs) relax the linearity assumption, and allow the data to determine
the fit of the model while permitting missing observations and different
correlation structures. Therefore, GAMs present an excellent choice to
analyze non-linear longitudinal data in the context of biomedical
research. This paper summarizes the limitations of rm-ANOVA and LMEMs
and uses simulated data to visually show how both methods produce biased
estimates when used on non-linear data. We also present the basic theory
of GAMs, and using trends of oxygen saturation in tumors reported in the
biomedical literature, we simulate example longitudinal data (2
treatment groups, 10 subjects per group, 5 repeated measures for each
group) to demonstrate how these models can be computationally
implemented. We show that GAMs are able to produce estimates that are
consistent with the trends of biomedical non-linear data even in the
case when missing observations exist (with 40\% of the simulated
observations missing), allowing reliable inference from the data. To
make this work reproducible, the code and data used in this paper are
available at: \url{https://github.com/aimundo/GAMs-biomedical-research}.
\end{abstract}

@@

A \emph{post hoc} analysis is often used in conjunction with rm-ANOVA to
perform repeated comparisons to estimate a \emph{p-value}, which in turn
is used as a measure of significance. Although it is possible that a
\emph{post hoc} analysis of rm-ANOVA is able to find ``significant''
\emph{p-values}( \emph{p}\textless0.05) from non-linear data, the
validity of such metric is dependent on how adequate the model fits the
data. In other words, \emph{p-values} are valid only if the model and
the data have good agreement; if that is not the case, a ``Type III''
error (known as ``model misspecification'') occurs\citep{dennis2019}.
For example, model misspecification will occur when a model that is only
able to explain linear responses (such as rm-ANOVA) is fitted to data
that follows a quadratic trend, thereby causing the resulting
\emph{p-values} and parameter estimates to be invalid \citep{wang2019}.

@@

\begin{equation}
y_{ijt} = \beta_0+\beta_1 \times time_{t} +\beta_2 \times treatment_{j} +\beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{ijt}\\ 
(\#eq:linear-model)
\end{equation}

In this model \(y_{ijt}\) is the response for subject \(i\), in
treatment group \(j\) at time \(t\), which can be decomposed in a mean
value \(\beta_0\), \emph{fixed effects} of time (\(time_t\)), treatment
(\(treatment_j\)) and their interaction \(time_t*treatment_j\) which
have linear slopes given by \(\beta_1, \beta_2\) and \(\beta_3\),
respectively. Independent errors \(\varepsilon_{ijt}\) represent random
variation not explained by the \emph{fixed} effects, and are assumed to
be \(\sim N(0,\sigma^2)\) (independently and identically normally
distributed with mean zero and variance \(\sigma^2\)). In a biomedical
research context, suppose two treatments groups are used in a study
(e.g., ``placebo'' vs.~``novel drug'' or ``saline''
vs.~``chemotherapy''). Then, the group terms in Equation
@ref(eq:linear-model) can be written as below with \(treatment_j=0\)
representing the first treatment group (Group A) and \(treatment_j=1\)
representing the second treatment group (Group B). With this notation,
the linear model then can be expressed as

\begin{equation}
y_{ijt} = \begin{cases}
\beta_0 + \beta_1\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A}\\
\beta_0 + \beta_2+\beta_1 \times time_{t} +\beta_3 \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}\\
\end{cases}
(\#eq:ANOVA-by-group)
\end{equation}

@@

\begin{equation}
y_{ijt} = \begin{cases}
\beta_0 + \beta_1\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A}\\
\widetilde{\beta_{0}} + \widetilde{\beta_1} \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}\\
\end{cases}
(\#eq:ANOVA-lines)
\end{equation}

Presenting the model in this manner makes clear that when treating
different groups, an rm-ANOVA model is able to accommodate non-parallel
lines in each case (different intercepts and slopes per group). In other
words, the rm-ANOVA model ``expects'' a linear relationship between the
covariates and the response, this means that either presented as
Equation @ref(eq:linear-model), Equation @ref(eq:ANOVA-by-group) or
Equation @ref(eq:ANOVA-lines), an rm-ANOVA model is only able to
accommodate linear patterns in the data. If the data show non-linear
behavior, the rm-ANOVA model will approximate this behavior with
non-parallel lines.

@@

\begin{equation}
y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\mu_{ij} +\varepsilon_{ijt}\\ 
(\#eq:LMEM)
\end{equation}

When Equation @ref(eq:linear-model) and Equation @ref(eq:LMEM) are
compared, it is easily noticeable that LMEM and rm-ANOVA have the same
construction regarding the \emph{fixed effects} of time and treatment,
but that the LMEM incorporates an additional source of variation (the
term \(\mu_{ij}\)). This term \(\mu_{ij}\) is the one that corresponds
to the \emph{random effect}, accounting for variability in each subject
(subject\(_i\)) within each group (group\(_j\)). The \emph{random}
component can also be understood as used to model some ``noise'' in the
response, but that is intended to be analyzed and disentangled from the
``global noise'' term \(\varepsilon_{ijt}\) from Equation
@ref(eq:linear-model).

For example, if the blood concentration of the drug is measured in
certain subjects in the early hours of the morning while other subjects
are measured in the afternoon, it is possible that the difference in the
collection time introduces some ``noise'' in the data. As the name
suggests, this ``random'' variability needs to be modeled as a variable
rather than as a constant value. The \emph{random effect} \(\mu_{ij}\)
in Equation @ref(eq:LMEM) is assumed to be
\(\mu_{ij} \sim N(0,\sigma^2_\mu)\). In essence, the \emph{random
effect} in a LMEM enables to fit models with different slopes at the
subject-level\citep{pinheiro2006}. However, the expected linear
relationship of the covariates and the response in Equation
@ref(eq:linear-model) and in Equation @ref(eq:LMEM) is essentially the
same, representing a major limitation of LMEMs to fit a non-linear
response.

@@

To visually demonstrate the limitations of rm-ANOVA an LMEMs for
non-linear longitudinal data, this section presents a simulation
experiment of a normally distributed response of two groups of 10
subjects each. An rm-ANOVA model (Equation @ref(eq:linear-model)), and a
LMEM (Equation @ref(eq:LMEM)) are fitted to each group, using
\(\textsf{R}\) \citep{r} and the package \emph{nlme} \citep{nlme}.

Briefly, two cases for the mean responses for each group are considered:
in the first case, the mean response in each group is a linear function
over time with different intercepts and slopes; a negative slope is used
for Group 1 and a positive slope is used for Group 2 (Figure
@ref(fig:l-q-response)A). In the second case, a second-degree polynomial
(quadratic) function is used for the mean response per group: the
quadratic function is concave down for Group 1 and it is concave up for
Group 2 (Figure @ref(fig:l-q-response)C). In both the linear and
quadratic simulated data, the groups start with the same mean value at
the first time point. This is intentional in order to simulate the
expected temporal evolution of some physiological quantity, which is
typical in biomedical experiments where a strong non-linear trend is
present.

Specifically, the rationale for the chosen linear and quadratic
functions is the expectation that a measured response in two treatment
groups is similar in the initial phase of the study, but as therapy
progresses a divergence in the trend of the response indicates a
treatment effect. In other words, Group 1 can be thought as a
``Control'' group and Group 2 as a ``Treatment'' group. From the mean
response per group (linear or quadratic), the variability or ``error''
of individual responses within each group is simulated using a
covariance matrix with compound symmetry (constant variance across
time). Thus, the response per subject in both the linear and quadratic
simulation corresponds to the mean response per group plus the error
(Figure @ref(fig:l-q-response) B,D).

A more comprehensive exploration of the fit of rm-ANOVA and LMEMs for
linear and non-linear longitudinal appears in Figure
@ref(fig:linear-cases-Appendix) and Figure
@ref(fig:quadratic-cases-Appendix) in the Appendix, where simulation
with compound symmetry and independent errors (errors generated from a
normal distribution that are not constant over time) and the plot of
simulated errors, and fitted parameters in presented. We are aware that
the simulated data used in this section present an extreme case that
might not occur frequently in biomedical research, but they are used as
a representation of the consequences of modeling non-linear data with a
linear model such as rm-ANOVA or LMEMs. Of notice, in Section
@ref(longitudinal-GAMs) we use simulated data that does follow reported
trends in the biomedical literature to implement GAMs.

(ref:l-q-response-caption) Simulated responses from two groups with
correlated errors using a LMEM and a rm-ANOVA model. Top row: linear
response, bottom row: quadratic response. A: Simulated linear data with
known mean response (thin lines) and individual responses (points)
showing the dispersion of the data. D: Simulated quadratic data with
known mean response (thin lines) and individual responses (points)
showing the dispersion of the data. B,E: Estimates from the rm-ANOVA
model for the mean group response (linear of quadratic). Points
represent the original raw data. The rm-ANOVA model not only fails to
pick the trend of the quadratic data (D) but also assigns a global
estimate that does not take between-subject variation. C, F: Estimates
from the LMEM in the linear and quadratic case. The LMEM incorporates a
random effect for each subject, but this model and the rm-ANOVA model
are unable to follow the trend of the data and grossly bias the initial
estimates for each group in the quadratic case (bottom row).

@@

\caption{(ref:l-q-response-caption)}\label{fig:l-q-response}
\end{figure}

The simulation shows that the fits produced by the LMEM and the rm-ANOVA
model are good for linear data, as the predictions for the mean response
are reasonably close to the ``truth'' of the simulated data (Figure
@ref(fig:l-q-response)A). When the linearity and compound symmetry
assumptions are met, the rm-ANOVA model approximates well the global
trend by group (Figure @ref(fig:l-q-response)B). Note that because the
LMEM incorporates \emph{random effects}, is able to provide estimates
for each subject and a ``global'' estimate (Figure
@ref(fig:l-q-response)C).

However, consider the case when the data follows a non-linear trend,
such as the simulated data in Figure @ref(fig:l-q-response)D. Here, the
mean response per group was simulated using a quadratic function, and
errors and individual responses were produced as in Figure
@ref(fig:l-q-response)A. The mean response in the simulated data with
quadratic behavior changes in each group through the timeline, and the
mean value is the same as the initial value by the fifth time point for
each group. Fitting an rm-ANOVA model (Equation @ref(eq:linear-model))
or a LMEM (Equation @ref(eq:LMEM)) to this data produces the fit that
appears in Figure @ref(fig:l-q-response)E, F.

Comparing the fitted responses of the LMEM and the rm-ANOVA models used
in the simulated quadratic data (Figure @ref(fig:l-q-response)E, F)
indicates that the models are not capturing the changes within each
group. Specifically, note that the fitted mean response of both models
shows that the change (increase for Treatment 1 or decrease for
Treatment 2) in the response through time points 2 and 4 is not being
captured. The LMEM is only able to account for between-subject variation
by providing estimates for each subject (Figure
@ref(fig:l-q-response)F), but both models are unable to capture the fact
that the initial values are the same in each group, and instead fit
non-parallel lines that have initial values that are markedly different
from the ``true'' initial values in each case (compare Figure
@ref(fig:l-q-response)D with Figure @ref(fig:l-q-response)E, F). If such
a change has important physiological implications, both rm-ANOVA and
LMEMs omit it from the fitted mean response. Thus, even though the model
correctly detects a divergence between treatment groups, the exact
nature of this difference is not correctly identified, limiting valuable
inferences from the data.

@@

\begin{equation}
  y_{ijt}=\beta_0+f(x_t\mid \beta_j)+\varepsilon_{ijt}
  (\#eq:GAM)
\end{equation}

@@

The smooth functional relationship between the covariates and the
response in GAMs is specified using a semi-parametric relationship that
can be fit within the GLM framework, by using \emph{basis function}
expansions of the covariates and by estimating random coefficients
associated with these basis functions. A \emph{basis} is a set of
functions that spans the mathematical space where the smooths that
approximate \(f(x_t\mid \beta_j)\) exist \citep{simpson2018}. For the
linear model in Equation @ref(eq:linear-model), the basis coefficients
are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are
\(time_t\), \(treatment_j\) and \(time_t \times treatment_j\). The basis
function then, is the combination of basis coefficients and basis
vectors that map the possible relationship between the covariates and
the response \citep{hefley2017}, which in the case of Equation
@ref(eq:linear-model) is restricted to a linear family of functions. In
the case of Equation @ref(eq:GAM), the basis functions are contained in
the expression \(f(x_t\mid \beta_j)\), which means that the model allows
for non-linear relationships among the covariates.

@@

To further clarify the concept of basis functions and smooth functions,
consider the simulated response for Group 1 in Figure
@ref(fig:l-q-response)C. The simplest GAM model that can be used to
estimate such response is that of a single smooth term for the time
effect; i.e., a model that fits a smooth to the trend of the group
through time. The timeline can be divided in equally spaced
\emph{knots}, each knot being a region where a different set of basis
functions will be used. Because there are six timepoints for this group,
five knots can be used. The model with five knots to construct the
smooth term means that it will have four basis functions (plus one that
corresponds to the intercept). The choice of basis functions is set
using default values in the package \emph{mgcv} depending on the number
of knots. In Figure @ref(fig:basis-plot)A, the four basis functions (and
the intercept) are shown. Each of the basis functions is composed of six
different points (because there are six points on the timeline). To
control the ``wiggliness'' of the fit, each of the basis functions of
Figure @ref(fig:basis-plot)A is weighted by multiplying it by a
coefficient according to the matrix of Figure @ref(fig:basis-plot)B. The
parameter estimates are penalized (shrunk towards 0) where the penalty
reduces the ``wiggliness'' of the smooth fit to prevent overfitting. A
weak penalty estimate will result in wiggly functions whereas a strong
penalty estimate provides evidence that a linear response is
appropriate.

To get the weighted basis functions, each basis (from Figure Figure
@ref(fig:basis-plot)A) is multiplied by the corresponding coefficients
in Figure @ref(fig:basis-plot)B, thereby increasing or decreasing the
original basis functions. Figure @ref(fig:basis-plot)C shows the
resulting weighted basis functions. Note that the magnitude of the
weighting for the first basis function has resulted in a decrease of its
overall value (because the coefficient for that basis function is less
than 1). On the other hand, the third basis function has roughly doubled
its value. Finally, the weighted basis functions are added at each
timepoint to produce the smooth term. The resulting smooth term for the
effect of \emph{time} is shown in Figure @ref(fig:basis-plot)D (orange
line), along the simulated values per group, which appear as points.

(ref:basis-plot-caption) Basis functions for a single smoother for time
with five knots. A: Basis functions for a single smoother for time for
the simulated data of Group 1 from Figure 2. B: Matrix of basis function
weights. Each basis function is multiplied by a coefficient which can be
positive or negative. The coefficient determines the overall effect of
each basis in the final smoother. C: Weighted basis functions. Each of
the four basis functions of panel A has been weighted by the
corresponding coefficient shown in Panel B. Note the corresponding
increase (or decrease) in magnitude of each weighted basis function. D:
Smoother for time and original data points. The smoother (line) is the
result of the sum of each weighted basis function at each time point,
with simulated values for the group shown as points.

@@

\caption{(ref:basis-plot-caption)}\label{fig:basis-plot}
\end{figure}

@@

The simulated data is based on the reported longitudinal changes in
oxygen saturation (\(\mbox{StO}_2\)) in subcutaneous tumors that appear
in Figure 3C in \citep{vishwanath2009}. In the paper, diffuse
reflectance spectroscopy was used to quantify \(\mbox{StO}_2\) changes
in both groups at the same time points (days 0, 2, 5, 7 and 10). In the
``Treatment'' group (chemotherapy) an increase in \(\mbox{StO}_2\) is
observed through time, while a decrease is seen in the ``Control''
(saline) group. Following the reported trend, we simulated 10 normally
distributed observations at each time point with a standard deviation
(SD) of 10\% (matching the SD in the original paper). The simulated and
real data appear in Figure @ref(fig:sim-smooth-plot)A and the inset,
respectively.

@@

\caption{(ref:pairwise-comp)}\label{fig:plot-pairwise-comp}
\end{figure}

@@

\hypertarget{bibliography}{%
\section{Bibliography}\label{bibliography}}

\hypertarget{appendix}{%
\section{(APPENDIX)}\label{appendix}}

>>>>

\usepackage{float}
\usepackage{listings}
\setcounter{secnumdepth}{2} %to add section numbers
\usepackage{chngcntr} %package to enable figure numbers by sections
%\counterwithin{figure}{section} %command for chngcntr in order to provide numbers per section that match with the letter for the Appendix

%to have page numbers with letters in the Appendix:
%solution from https://tex.stackexchange.com/questions/59572/custom-page-numbering-for-appendix
\newcommand{\appendixpagenumbering}{
  \break
  \pagenumbering{arabic}
  \renewcommand{\thepage}{\thesection-\arabic{page}}
}

%newcommand{\passthrough}[1]{#1}

%theme colors for the code chunks (originally from latex-solarized on GitHub)
%https://github.com/jez/latex-solarized
 \usepackage{xcolor}
\definecolor{sbase03}{HTML}{002B36}
\definecolor{sbase02}{HTML}{073642}
\definecolor{sbase01}{HTML}{586E75}
\definecolor{sbase00}{HTML}{657B83}
\definecolor{sbase0}{HTML}{839496}
\definecolor{sbase1}{HTML}{93A1A1}
\definecolor{sbase2}{HTML}{EEE8D5}
\definecolor{sbase3}{HTML}{FDF6E3}
\definecolor{syellow}{HTML}{B58900}
\definecolor{sorange}{HTML}{CB4B16}
\definecolor{sred}{HTML}{DC322F}
\definecolor{smagenta}{HTML}{D33682}
\definecolor{sviolet}{HTML}{6C71C4}
\definecolor{sblue}{HTML}{268BD2}
\definecolor{scyan}{HTML}{2AA198}
\definecolor{sgreen}{HTML}{859900}
%command to set parameter(s) in package listings
\lstdefinestyle{astyle}{
    % How/what to match
    language=R,
    sensitive=true,
    % Border (above and below)
    frame=lines,
    % Extra margin on line (align with paragraph)
    xleftmargin=\parindent,
    % Put extra space under caption
    belowcaptionskip=1\baselineskip,
    % Colors
    backgroundcolor=\color{sbase3},
    basicstyle=\color{sbase00}\ttfamily,
    keywordstyle=\color{scyan},
    commentstyle=\color{sbase1},
    stringstyle=\color{sblue},
    numberstyle=\color{sviolet},
    identifierstyle=\color{sbase00},
    % Break long lines into multiple lines?
    breaklines=true,
    % Show a character for spaces?
    showstringspaces=false,
    tabsize=2
}

\lstset{style=astyle}

@@

\begin{abstract}
In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the \emph{repeated measures analysis of variance} (rm-ANOVA) or more recently, \emph{linear mixed models} (LMEMs). Although LMEMs are less restrictive than rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear, which is a common occurrence in biomedical research. In contrast, \emph{generalized additive models} (GAMs) relax the linearity assumption, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures. Therefore, GAMs present an excellent choice to analyze non-linear longitudinal data in the context of biomedical research. This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data. We also present the basic theory of GAMs, and using trends of oxygen saturation in tumors reported in the biomedical literature, we simulate example longitudinal data (two treatment groups, 10 subjects per group, five repeated measures for each group) to demonstrate how these models can be computationally implemented. We show that GAMs are able to produce estimates that are consistent with the trends of biomedical non-linear data even in the case when missing observations exist (with 40\% of the simulated observations missing), allowing reliable inference from the data. To make this work reproducible, the code and data used in this paper are available at: \url{https://github.com/aimundo/GAMs-biomedical-research}.
\end{abstract}

@@

A \emph{post hoc} analysis is often used in conjunction with rm-ANOVA to perform repeated comparisons to estimate a \emph{p-value}, which in turn is used as a measure of significance.
Although it is possible that a \emph{post hoc} analysis of rm-ANOVA is able to find ``significant'' \emph{p-values}( \emph{p}\textless{}0.05) from non-linear data, the validity of such metric is dependent on how adequate the model fits the data. In other words, \emph{p-values} are valid only if the model and the data have good agreement; if that is not the case, a ``Type III'' error (known as ``model misspecification'') occurs\citep{dennis2019}. For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting \emph{p-values} and parameter estimates to be invalid \citep{wang2019}.

@@

\begin{equation}
y_{ijt} = \beta_0+\beta_1 \times time_{t} +\beta_2 \times treatment_{j} +\beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{ijt}\\ 
\label{eq:linear-model}
\end{equation}

In this model \(y_{ijt}\) is the response for subject \(i\), in treatment group \(j\) at time \(t\), which can be decomposed in a mean value \(\beta_0\), \emph{fixed effects} of time (\(time_t\)), treatment (\(treatment_j\)) and their interaction \(time_t*treatment_j\) which have linear slopes given by \(\beta_1, \beta_2\) and \(\beta_3\), respectively. Independent errors \(\varepsilon_{ijt}\) represent random variation not explained by the \emph{fixed} effects, and are assumed to be \(\sim N(0,\sigma^2)\) (independently and identically normally distributed with mean zero and variance \(\sigma^2\)).
In a biomedical research context, suppose two treatments groups are used in a study (e.g., ``placebo'' vs.~``novel drug'' or ``saline'' vs.~``chemotherapy''). Then, the group terms in Equation \eqref{eq:linear-model} can be written as below with \(treatment_j=0\) representing the first treatment group (Group A) and \(treatment_j=1\) representing the second treatment group (Group B). With this notation, the linear model then can be expressed as

\begin{equation}
y_{ijt} = \begin{cases}
\beta_0 + \beta_1\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A}\\
\beta_0 + \beta_2+\beta_1 \times time_{t} +\beta_3 \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}\\
\end{cases}
\label{eq:ANOVA-by-group}
\end{equation}

@@

\begin{equation}
y_{ijt} = \begin{cases}
\beta_0 + \beta_1\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A}\\
\widetilde{\beta_{0}} + \widetilde{\beta_1} \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}\\
\end{cases}
\label{eq:ANOVA-lines}
\end{equation}

Presenting the model in this manner makes clear that when treating different groups, an rm-ANOVA model is able to accommodate non-parallel lines in each case (different intercepts and slopes per group). In other words, the rm-ANOVA model ``expects'' a linear relationship between the covariates and the response, this means that either presented as Equation \eqref{eq:linear-model}, Equation \eqref{eq:ANOVA-by-group} or Equation \eqref{eq:ANOVA-lines}, an rm-ANOVA model is only able to accommodate linear patterns in the data. If the data show non-linear behavior, the rm-ANOVA model will approximate this behavior with non-parallel lines.

@@

\begin{equation}
y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\mu_{ij} +\varepsilon_{ijt}\\ 
\label{eq:LMEM}
\end{equation}

When Equation \eqref{eq:linear-model} and Equation \eqref{eq:LMEM} are compared, it is easily noticeable that LMEM and rm-ANOVA have the same construction regarding the \emph{fixed effects} of time and treatment, but that the LMEM incorporates an additional source of variation (the term \(\mu_{ij}\)). This term \(\mu_{ij}\) is the one that corresponds to the \emph{random effect}, accounting for variability in each subject (subject\(_i\)) within each group (group\(_j\)). The \emph{random} component can also be understood as used to model some ``noise'' in the response, but that is intended to be analyzed and disentangled from the ``global noise'' term \(\varepsilon_{ijt}\) from Equation \eqref{eq:linear-model}.

For example, if the blood concentration of the drug is measured in certain subjects in the early hours of the morning while other subjects are measured in the afternoon, it is possible that the difference in the collection time introduces some ``noise'' in the data. As the name suggests, this ``random'' variability needs to be modeled as a variable rather than as a constant value. The \emph{random effect} \(\mu_{ij}\) in Equation \eqref{eq:LMEM} is assumed to be \(\mu_{ij} \sim N(0,\sigma^2_\mu)\). In essence, the \emph{random effect} in a LMEM enables to fit models with different slopes at the subject-level\citep{pinheiro2006}. However, the expected linear relationship of the covariates and the response in Equation \eqref{eq:linear-model} and in Equation \eqref{eq:LMEM} is essentially the same, representing a major limitation of LMEMs to fit a non-linear response.

@@

To visually demonstrate the limitations of rm-ANOVA an LMEMs for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each. An rm-ANOVA model (Equation \eqref{eq:linear-model}), and a LMEM (Equation \eqref{eq:LMEM}) are fitted to each group, using \(\textsf{R}\) \citep{r} and the package \emph{nlme} \citep{nlme}.

Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function over time with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \ref{fig:l-q-response}A). In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \ref{fig:l-q-response}C). In both the linear and quadratic simulated data, the groups start with the same mean value at the first time point. This is intentional in order to simulate the expected temporal evolution of some physiological quantity, which is typical in biomedical experiments where a strong non-linear trend is present.

Specifically, the rationale for the chosen linear and quadratic functions is the expectation that a measured response in two treatment groups is similar in the initial phase of the study, but as therapy progresses a divergence in the trend of the response indicates a treatment effect. In other words, Group 1 can be thought as a ``Control'' group and Group 2 as a ``Treatment'' group. From the mean response per group (linear or quadratic), the variability or ``error'' of individual responses within each group is simulated using a covariance matrix with compound symmetry (constant variance across time). Thus, the response per subject in both the linear and quadratic simulation corresponds to the mean response per group plus the error (Figure \ref{fig:l-q-response} B,D).

A more comprehensive exploration of the fit of rm-ANOVA and LMEMs for linear and non-linear longitudinal appears in Figure \ref{fig:linear-cases-Appendix} and Figure \ref{fig:quadratic-cases-Appendix} in the Appendix, where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented. We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used as a representation of the consequences of modeling non-linear data with a linear model such as rm-ANOVA or LMEMs. Of notice, in Section \ref{longitudinal-GAMs} we use simulated data that does follow reported trends in the biomedical literature to implement GAMs.

@@

\caption{Simulated responses from two groups with correlated errors using a LMEM and a rm-ANOVA model. Top row: linear response, bottom row: quadratic response. A: Simulated linear data with known mean response (thin lines) and individual responses (points) showing the dispersion of the data. D: Simulated quadratic data with known mean response (thin lines) and individual responses (points) showing the dispersion of the data. B,E: Estimates from the rm-ANOVA model for the mean group response (linear of quadratic). Points represent the original raw data. The rm-ANOVA model not only fails to pick the trend of the quadratic data (D) but also assigns a global estimate that does not take between-subject variation. C, F: Estimates from the LMEM in the linear and quadratic case. The LMEM incorporates a random effect for each subject, but this model and the rm-ANOVA model are unable to follow the trend of the data and grossly bias the initial estimates for each group in the quadratic case (bottom row).}\label{fig:l-q-response}
\end{figure}

The simulation shows that the fits produced by the LMEM and the rm-ANOVA model are good for linear data, as the predictions for the mean response are reasonably close to the ``truth'' of the simulated data (Figure \ref{fig:l-q-response}A). When the linearity and compound symmetry assumptions are met, the rm-ANOVA model approximates well the global trend by group (Figure \ref{fig:l-q-response}B). Note that because the LMEM incorporates \emph{random effects}, is able to provide estimates for each subject and a ``global'' estimate (Figure \ref{fig:l-q-response}C).

However, consider the case when the data follows a non-linear trend, such as the simulated data in Figure \ref{fig:l-q-response}D. Here, the mean response per group was simulated using a quadratic function, and errors and individual responses were produced as in Figure \ref{fig:l-q-response}A. The mean response in the simulated data with quadratic behavior changes in each group through the timeline, and the mean value is the same as the initial value by the fifth time point for each group. Fitting an rm-ANOVA model (Equation \eqref{eq:linear-model}) or a LMEM (Equation \eqref{eq:LMEM}) to this data produces the fit that appears in Figure \ref{fig:l-q-response}E, F.

Comparing the fitted responses of the LMEM and the rm-ANOVA models used in the simulated quadratic data (Figure \ref{fig:l-q-response}E, F) indicates that the models are not capturing the changes within each group. Specifically, note that the fitted mean response of both models shows that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured. The LMEM is only able to account for between-subject variation by providing estimates for each subject (Figure \ref{fig:l-q-response}F), but both models are unable to capture the fact that the initial values are the same in each group, and instead fit non-parallel lines that have initial values that are markedly different from the ``true'' initial values in each case (compare Figure \ref{fig:l-q-response}D with Figure \ref{fig:l-q-response}E, F). If such a change has important physiological implications, both rm-ANOVA and LMEMs omit it from the fitted mean response. Thus, even though the model correctly detects a divergence between treatment groups, the exact nature of this difference is not correctly identified, limiting valuable inferences from the data.

@@

\begin{equation}
  y_{ijt}=\beta_0+f(x_t\mid \beta_j)+\varepsilon_{ijt}
  \label{eq:GAM}
\end{equation}

@@

The smooth functional relationship between the covariates and the response in GAMs is specified using a semi-parametric relationship that can be fit within the GLM framework, by using \emph{basis function} expansions of the covariates and by estimating random coefficients associated with these basis functions. A \emph{basis} is a set of functions that spans the mathematical space where the smooths that approximate \(f(x_t\mid \beta_j)\) exist \citep{simpson2018}. For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(time_t\), \(treatment_j\) and \(time_t \times treatment_j\). The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response \citep{hefley2017}, which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions. In the case of Equation \eqref{eq:GAM}, the basis functions are contained in the expression \(f(x_t\mid \beta_j)\), which means that the model allows for non-linear relationships among the covariates.

@@

To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure \ref{fig:l-q-response}C. The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time. The timeline can be divided in equally spaced \emph{knots}, each knot being a region where a different set of basis functions will be used. Because there are six timepoints for this group, five knots can be used. The model with five knots to construct the smooth term means that it will have four basis functions (plus one that corresponds to the intercept). The choice of basis functions is set using default values in the package \emph{mgcv} depending on the number of knots. In Figure \ref{fig:basis-plot}A, the four basis functions (and the intercept) are shown. Each of the basis functions is composed of six different points (because there are six points on the timeline). To control the ``wiggliness'' of the fit, each of the basis functions of Figure \ref{fig:basis-plot}A is weighted by multiplying it by a coefficient according to the matrix of Figure \ref{fig:basis-plot}B. The parameter estimates are penalized (shrunk towards 0) where the penalty reduces the ``wiggliness'' of the smooth fit to prevent overfitting. A weak penalty estimate will result in wiggly functions whereas a strong penalty estimate provides evidence that a linear response is appropriate.

To get the weighted basis functions, each basis (from Figure Figure \ref{fig:basis-plot}A) is multiplied by the corresponding coefficients in Figure \ref{fig:basis-plot}B, thereby increasing or decreasing the original basis functions. Figure \ref{fig:basis-plot}C shows the resulting weighted basis functions. Note that the magnitude of the weighting for the first basis function has resulted in a decrease of its overall value (because the coefficient for that basis function is less than 1). On the other hand, the third basis function has roughly doubled its value. Finally, the weighted basis functions are added at each timepoint to produce the smooth term. The resulting smooth term for the effect of \emph{time} is shown in Figure \ref{fig:basis-plot}D (orange line), along the simulated values per group, which appear as points.

@@

\caption{Basis functions for a single smoother for time with five knots. A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2. B: Matrix of basis function weights. Each basis function is multiplied by a coefficient which can be positive or negative. The coefficient determines the overall effect of each basis in the final smoother. C: Weighted basis functions. Each of the four basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B. Note the corresponding increase (or decrease) in magnitude of each weighted basis function. D: Smoother for time and original data points. The smoother (line) is the result of the sum of each weighted basis function at each time point, with simulated values for the group shown as points.}\label{fig:basis-plot}
\end{figure}

@@

The simulated data is based on the reported longitudinal changes in oxygen saturation (\(\mbox{StO}_2\)) in subcutaneous tumors that appear in Figure 3C in \citep{vishwanath2009}. In the paper, diffuse reflectance spectroscopy was used to quantify \(\mbox{StO}_2\) changes in both groups at the same time points (days 0, 2, 5, 7 and 10). In the ``Treatment'' group (chemotherapy) an increase in \(\mbox{StO}_2\) is observed through time, while a decrease is seen in the ``Control'' (saline) group. Following the reported trend, we simulated 10 normally distributed observations at each time point with a standard deviation (SD) of 10\% (matching the SD in the original paper).
The simulated and real data appear in Figure \ref{fig:sim-smooth-plot}A and the inset, respectively.

@@

\caption{Pairwise comparisons for smooth terms. A: Pairwise comparisons for the full dataset. B: Pairwise comparisons for the dataset with missing observations. Significant differences exist where the 95\% empirical Bayesian credible interval does not cover 0. In both cases the effect of treatment is significant after day 3.}\label{fig:plot-pairwise-comp}
\end{figure}

@@

\hypertarget{appendix-appendix}{%
\appendix}

\appendixpagenumbering

\counterwithin{figure}{section}

\hypertarget{appendix}{%
\section{APPENDIX}\label{appendix}}

This section presents the code used to generate figures, models and simulated data from the main manuscript.
